---
title: "Baseball Analysis"
output:
  word_document: default
  html_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE}
if(!require(psych)){
  install.packages("psych")
  library(psych)
}
if(!require(flextable)){
  install.packages("flextable")
  library(flextable)
}

if(!require(corrplot)){
  install.packages("corrplot")
  library(corrplot)
}

if(!require(ggplot2)){
  install.packages("ggplot2")
  library(ggplot2)
}

if(!require(easystats)){
  install.packages("easystats")
  library(easystats)
}

```


# Part 1 - Accessing and Verifying data

Step 1:
Load the data

```{r}

BaseBall<-read.csv("../data/baseball.csv",sep=",",header=TRUE)
```

Check the dimensions
```{r}
dim(BaseBall)
```


Check the structure

```{r}
str(BaseBall)
```

List out variable names

```{r}
colnames(BaseBall)
```


Have a think! Perhaps we could load the data in a different way where it may be more useful, by changing some options!


```{r}
baseball<-read.csv("../data/baseball.csv",sep=",",header=TRUE,row.names = 'Name',stringsAsFactors = TRUE)

```

```{r}
str(baseball)
```

Here we have told R to load our dataset, with every string treated as if it was a categorical variable (not always a good idea!) and that the column called 'Name' should be used as a row name, rather than a useful variable.


Which of the two datasets should we use to proceed?
BaseBall or baseball?

Or should we make further modifications?

```{r}
rm(BaseBall)
```

# Part 2 - Some initial explorations

1. Check for NAs or NULLs!

```{r}
sum(is.na(baseball))
sum(is.null(baseball))

```

```{r}
length(unique(rownames(baseball)))

```


```{r}
summary(baseball)
```

```{r}
Sal_Div<-describeBy(baseball$Salary,group = baseball$Division,mat=TRUE)
Sal_Div<-Sal_Div[,c("group1","n","mean","median","sd","min","max")]
colnames(Sal_Div)[1]<-"division"
Sal_Div$mean<-signif(Sal_Div$mean ,4)
flextable(Sal_Div)|>theme_booktabs(x,bold_header = TRUE)
```
## Find numeric variables only

```{r}
baseball_num<-Filter(is.numeric,baseball)

```

```{r}
X<-describe(baseball_num)
X$vars<-rownames(X)
flextable(X[,c("vars","n","mean","median","sd","min","max")])|>theme_booktabs(x,bold_header = TRUE)

```

```{r}
M<-cor(baseball_num)
corrplot(M,method="circle",type="lower",diag=FALSE,tl.col = "black",tl.srt = 45)
```

```{r}
which.max(abs(M["Salary",-17]))

```

```{r}
ggplot(data=baseball,aes(x=CRBI,y=Salary,colour=Division,shape=Division))+geom_point()+theme_minimal()

```

# Part 3 - Hypothesis testing

```{r}
TTest<-t.test(baseball$Salary~baseball$Division,paired=FALSE)
```


```{r}
report(TTest)
```
The Welch Two Sample t-test testing the difference of
Salary by Division (mean in group E =
624.27, mean in group W = 450.88) suggests that the
effect is positive, statistically significant
(difference = 173.39, 95% CI [64.73, 282.06], t(218)
= 3.14, p = 0.002.

```{r}
CTest<-cor.test(baseball$Salary,baseball$CRBI)
report(CTest)
```

The Pearson's product-moment correlation between
Salary and CRBI is positive,
statistically significant (r = 0.57, 95%
CI [0.48, 0.64], df=261, p < .001)

# Part 4 - Constructing a simple linear regression

```{r}
mod1<-lm(Salary~CRBI,data=baseball)
as_flextable(mod1)

report(mod1)
```

# Part 5 - Constructing a multiple linear regression

```{r}
mod_full<-lm(Salary~.,data=baseball)
summary(mod_full)

```

```{r}
check_model(mod_full,check=c("vif","ncv"))

```

```{r}
mod_red<-step(mod_full,verbose=FALSE)
summary(mod_red)
```

```{r}
mod_redB<-step(mod_full,k=log(nrow(baseball)))
summary(mod_redB)
```

```{r}
check_model(mod_redB,check=c("vif","linearity"))
```


# Part 6 - Reporting your results


- Is the data relevant / appropriate!

The data analysed was collected in the 1980s (30+ years ago). Therefore the relevance of the variables is questionable and the magnitude of the coefficient estimates do not take factors such as inflation into account.


We have many concerns about the final model selected.  There are still variables included that result in high multicollinearity; this means that our independent variables are not independent of one another. It causes problems with interpretation (cannot assume other variables remain constant); it can also cause problems with the numerical stability of algorithms which in term can affect model estimates of coefficients, standard errors of coefficients and hence associated p-values. 
Therefore we cannot be certain about claims that particular variables may or may not be statistically significant.

We also observe that our residuals are not constant. This indicates that we may need to apply a transformation to the data in order to ensure that assumptions of a linear model become appropriate (residuals are normally distributed with a mean of zero and a constant standard deviation; the residuals should also be independent of one another)

- Do we have concerns about the final model
  - in terms of model assumptions
  - in terms of interpretability

Key findings: what variables are important
what variables aren't important!





