---
title: "Metadata and Exploratory Data Analysis"
author: "Statistical Inference"
date: "17/03/2026"
output: word_document
---

# Preparing this worksheet

### Exercise 1: Install packages

Make sure you have the following packages installed, as they will be used in this worksheet:

- dplyr
- RSQLite
- assertr

```{r setup, include=FALSE}
library(dplyr)
library(RSQLite)
library(assertr)
knitr::opts_chunk$set(echo = TRUE)
```

### Exercise 2: Check data will load correctly

This worksheet uses the file `hsfe.db`

- You are expected to save the file `hsfe.db` in your `data` subfolder.
- You are expected to save this worksheet in your `scripts` subfolder.
- This is best practice for reproducible research in Statistical Inference.

Run the following code chunk to check if you can load the data for this worksheet.

```{r check_folder_structure, echo=FALSE}
data_location <- "../data/hsfe.db"
print(paste("My working directory is currently", getwd()))
print(paste("R is looking for the data in file", normalizePath(data_location)))
print(paste("R has", c("not")[!file.exists(data_location)] ,"found the data for this worksheet"))
if(!file.exists(data_location)) stop("R has not found the data for this worksheet") # Stops knitting if data not present

# If it does not work then:
# Make sure hsfe.db is saved in your data subfolder
# Make sure this worksheet is saved in your scripts subfolder
# Make sure your working directory is your scripts subfolder
# Make sure your knit directory is your scripts subfolder
```

# Health Survey for England Teaching Dataset

## Metadata about this dataset

The Dublin Core are a set of standards for creating and publishing metadata.
These standards recommend at least 15 core metadata elements are recorded for a dataset.

### Exercise 3: Record metadata

Complete the 15 Dublin Core metadata elements below for the Health Survey for England Teaching Dataset.
Some have already been filled in - you may wish to verify them.
You can find the information you need at https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=7402

- Title: (A name given to the dataset)
- Subject: (The topic of the dataset)
- Description: A teaching dataset based on the Health Survey for England
- Creator: (An entity primarily responsible for making the dataset)
- Publisher: (An entity responsible for making the dataset available)
- Contributor: None
- Date: 10 October 2013
- Type: Dataset
- Format: .db with 10,617 rows
- Identifier: https://beta.ukdataservice.ac.uk/datacatalogue/studies/study?id=7402
- Source: Health Survey for England, 2011
- Language: English
- Relation: User guide: https://doc.ukdataservice.ac.uk/doc/7402/mrdoc/pdf/7402hse_2011_teaching_dataset_user_guide.pdf
- Coverage: (The spatial and temporal topic of the dataset)
- Rights: Crown copyright. Safeguarded by UK data service (https://ukdataservice.ac.uk/find-data/access-conditions/)

For more about the Dublin Core see http://www.dublincore.org or the preparatory reading at https://guides.library.ucsc.edu/c.php?g=618773


### Exercise 4: Find out about variables using metadata   

Find out what the following variable names represent.
Use the user guide to help you - this is linked in the metadata.

- `Age` 
- `Sex` 
- `omsysval` 
- `porfv` 
- `econact`


## Loading data from a database using RSQLite

The R package `RSQLite` allows us to load data from a `.db` file using SQL comments (seen in other modules).

### Obtaining the names of data tables in the database

A database may contain many tables that contain data.
Unfortunately the metadata does not tell us which table to use. 
The following code chunk makes a connection to the database and returns the name of all data tables

```{r}
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(), data_location)
tabs <- dbListTables(con)
dbDisconnect(con)
tabs
```

There is only one table, so we assume that all data is contained within this table. 
We should make a note of this in our metadata, for future reference.


### Load data for this example

We are going to load a data frame containing just those variables listed in the exercise above. 
The following code chunk makes a connection to the database, runs a query to obtain the variables, then loads them into a data frame.

```{r}
library(RSQLite)
con <- dbConnect(RSQLite::SQLite(), data_location)
res <- dbSendQuery(con, "SELECT Age, Sex, omsysval, porfv, econact
                   FROM hse2011_teaching")
hse <- dbFetch(res)
dbClearResult(res)
dbDisconnect(con)

str(hse)
```

### Exercise 5: Loading data using RSQLite

Annotate the code chunk above: write a comment at the end of each line explaining what that line of code does.


## Preparing data using metadata and exploratory data analysis

The user guide linked within the metadata tells us that:

- `Age` is numeric and should have no missing values.
- `Sex` is coded as `1 = Male` and `2 = Female` and should have no missing values.
- `omsysval` is numeric but may have missing values.
- `porfv` is numeric but may have missing values.
- `econact` is coded as `1 = In employment`, `2 = Unemployed`, 3 = `Retired`, `4 = Other` with missing values `-9`, `-8` and `-1`.

We would want to verify this in the data we loaded, as part of our data preparation.
The following code chunk shows some ways to do this.

```{r}
str(hse)

# Check `Age` has no missing or unusual values
summary(hse$Age)
hist(hse$Age)

# Code `Sex` and check for missing values
hse$Sex <- factor(hse$Sex, levels=c(1,2), labels=c("Male", "Female"))
summary(hse$Sex)
sum(is.na(hse$Sex))

# Check 'omsysval' for missing or unusual values
summary(hse$omsysval)
# Blood pressure cannot be negative or zero, so recode these as missing
hse$omsysval[hse$omsysval <= 0] <- NA
summary(hse$omsysval)
hist(hse$omsysval)

# Check `porfv` for missing or unusual values
summary(hse$porfv)
# Portions cannot be negative, so recode these as missing
hse$porfv[hse$porfv < 0] <- NA
summary(hse$porfv)
hist(hse$porfv)

# Code `econact`
hse$econact[hse$econact == -9 | hse$econact == -8 | hse$econact == -1] <- NA
hse$econact <- factor(hse$econact, levels=c(1,2,3,4), labels=c("Employed","Unemployed","Retired","Other"))
summary(hse$econact)
```

## Preparing data using the assertR package

Data in external sources can change.
A dataset may have previously been prepared (cleaned and checked), but new data may contain new errors or values that will be unexpected later in the analysis pipeline.
We need to complete the process of cleaning and checking the data every time the dataset is updated.
If the dataset is updated regularly, we may want to automate this process.
One way to automatically check and clean updated data is using the assertR package.

```{r}
hse %>%
  verify(has_all_names("Age","Sex","omsysval","porfv","econact")) %>%
  assert(not_na, Age, Sex) %>%
  verify(Age >= 0) %>%
  verify(Age <= 120) %>%
  assert(in_set("Male","Female"), Sex) %>%
  assert(within_bounds(50,250), omsysval) %>%
  assert(within_bounds(0,30), porfv) %>%
  assert(in_set("Employed","Unemployed","Retired","Other"), econact) %>%
  summary()
```

### Exercise 6: Cleaning and checking using assertr

Annotate the code chunk above: write a comment at the end of each line explaining what that line of code does.


## Example exploratory data analysis

In this example, the research question is 'can we predict systolic blood pressure?'

```{r EDA}
str(hse)

# Univariate summaries
boxplot(hse$Age)
summary(hse$Age)

table(hse$Sex)

boxplot(hse$omsysval)
summary(hse$omsysval)

boxplot(hse$porfv)
summary(hse$porfv)

table(hse$econact)

# Comparing with the dependent variable
plot(hse$Age, hse$omsysval)
with(hse, cor(Age, omsysval, use="pairwise.complete.obs"))

boxplot(hse$omsysval ~ hse$Sex)

plot(hse$porfv, hse$omsysval)
with(hse, cor(porfv, omsysval, use="pairwise.complete.obs"))

boxplot(hse$omsysval ~ hse$econact)
```

### Exercise 7: Interpretation of EDA

(FILL IN XX VALUES, DELETE AS APPROPRIATE)

The univariate summaries tell us:

- The average age in the sample is approximately XX years. We see that 50% of the sample is aged between XX and XX years.
- The majority of the sample are MALE/FEMALE.
- Systolic blood pressure has a SYMMETRICAL/POSITIVELY SKEWED/NEGATIVELY SKEWED distribution with more outliers in the POSITIVE/NEGATIVE direction.
- Portions of fruit has a SYMMETRICAL/POSITIVELY SKEWED/NEGATIVELY SKEWED distribution with more outliers in the POSITIVE/NEGATIVE direction.
- The modal economic activity in the sample is EMPLOYED/UNEMPLOYED/RETIRED/OTHER.

The comparisons with the dependent variable tell us:

- There is a STRONG/MODERATE/MILD/NO POSITIVE/NEGATIVE correlation between age and systolic blood pressure.
- Blood pressure tends to be distributed higher in MALES/FEMALES.
- There is a STRONG/MODERATE/MILD/NO POSITIVE/NEGATIVE correlation between portions of fruit and systolic blood pressure.
- Blood pressure tends to be distributed higher in the EMPLOYED/UNEMPLOYED/RETIRED/OTHER category of economic activity.


### Exercise 7b: (for practice out of class)

Modify the EDA code chunk to improve the appearance of the plots (perhaps using the `ggplot2` package).


## Exercise 8: Meta-exercise

Look up the user guide for the Health Survey for England Teaching Dataset, and choose __at least one__ new variable that might be a predictor of systolic blood pressure.

Repeat this worksheet to perform a new analysis including this new variable. 
You will need to find out metadata for the new variable, including how it is coded.
You will need to modify the code chunks to:

- Load in this variable as well as the others.
- Check for missing or unusual values in the new variable, and
- Label the values for this variable, if necessary.
- Add in expectations for this variable, using assertr.
- Incorporate suitable exploratory data analysis designed to answer the research question.


