---
title: "Worksheet on multiple regression"
author: "Andrew Smith"
date: "`r Sys.Date()`"
output:
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```


# Multiple regression models

* We will use the 'mtcars' dataset that we saw when we studied simple linear regression. It is a built-in dataset in R: there is no need to load it.

- Dependent variable is 'mpg'
- Independent variables are drawn from all other variables in the dataset


## Reminder of simple linear regression model

In the last exercise of the simple linear regression worksheet, we looked for plots and linear correlations within the continuous variables of mtcars:

```{r}
df <- mtcars %>%                             # make data frame containing only variables we know to be continuous
  dplyr::select(-c(am, vs, cyl, gear, carb)) # need to use dplyr::select to distinguish from the function called 'select' in the MASS package

plot(df) # plot the continuous variables against each other

corrs <- cor(df) # calculate correlation coefficients

library(corrplot)
corrplot(corrs, method="number")

```

We found that `wt` had the strongest correlation with `mpg`, and fitted a simple linear regression model:

```{r}
model_1 <- lm(mpg ~ wt, data=mtcars)
summary(model_1)
```

* Notice R-squared = 0.7528. Means this model explains 75% of variation in mpg 
* Notice p-value for whole model (1.294e-10) equals the p-value for the coefficient of wt

```{r}
library(performance)
check_model(model_1)
```


## Model with two independent variables

Now try a model with two independent variables. Let's regress `mpg` on `wt` and `hp`

```{r}
model_2 <- lm(mpg ~ wt + hp, data=mtcars)
summary(model_2)
```

* Notice R-squared has gone up. This model with two independent variables explains 83% of variation in mpg
* Now we have a p-value for each coefficient, and they do not correspond to the p-value for the whole model (9.109e-12)

```{r}
check_model(model_2)
```


## Full model
What if we put in the full amount of independent variables?

* We would expect to be able to explain a large amount of variation in `mpg`, so why not put in all the independent variables we can!?
* If using all independent variables, would want to make sure we remove rows with NA values.

```{r}
mtcars <- na.omit(mtcars) # removes rows with NA values, if there are any

model_full <- lm(mpg ~ ., data=mtcars)
summary(model_full)
```

* All the independent variables together explain 87% of variation in mpg
* The p-value for the whole model is significant (3.793e-07) but none of the coefficients are significant!

```{r}
check_model(model_full)
```

There is a problem with multicollinearity!

* Multicollinearity is caused by correlation between the independent variables.
* We should be careful to check the correlation between the independent (predictor) variables.


# Dewis exercise on multiple regression

```{r dewis_multiple_regression}

### Use this code chunk for your code for the multiple regression Dewis exercise

```


# Akaike Information Criterion (AIC)

* We need to remove some independent variables from the full model
* Too many independent variables, and we get multicollinearity problems.
* Too many independent variables, we risk including an independent variable that makes no contribution to the model.
* Need a balance between 
  + Having enough independent variables to predict a decent proportion of variation in the dependent variable
  + Not having too many independent variables; not including independent variables that make no contribution.
* One way of achieving this balance is to __minimise the AIC__
  + Find out more about AIC in the help file `?extractAIC`

## Null model

Out of interest, fit a model with no independent variables

```{r}
model_null <- lm(mpg ~ 1, data=mtcars)
summary(model_null)
mean(mtcars$mpg)
```

* This model just fits the mean. It explains 0% of the variation in mpg

## Compare AICs

```{r}
library(MASS)
extractAIC(model_null)
extractAIC(model_1)
extractAIC(model_2)
extractAIC(model_full)
```

* It would appear that `model_2`, the model with `wt` and `hp`, is the best balanced model.
* But is there an even better model with a different number of independent variables?
* Use stepwise model selection to find the model with smallest AIC.

## Stepwise model selection

```{r}
# Tell stepAIC to start with the full model, and then save as the final model
model_final <- stepAIC(model_full)
summary(model_final)
```

* Final model explains XX% of variation in mpg

```{r}
check_model(model_final)
```

### Question
What is the interpretation of the final model?

One way to interpret is with a formula:

`mpg` = beta_0 + beta_1 * `wt` + beta_2 * `qsec` + beta_3 * `am`






# Advanced [optional]
# Bootstrap confidence intervals

Bootstrapping may be appropriate here because

* The residuals of the final model may not satisfy all of the assumptions
* Stepwise model selection can introduce bias to the usual statistical inference

## Find 95% confidence interval for coefficient of wt





```{r}
# Set up bootstrap
Nbootstrap<- 1000 # 1000 takes half a minute
coeff_wt <-rep(NA,Nbootstrap)

# Perform bootstrap
set.seed(4363)
for(i in seq_len(Nbootstrap)){
  usevalues<-sample(seq_along(mtcars$mpg),size=length(mtcars$mpg),replace=TRUE) 
  bootstrap.sample<- mtcars[usevalues,]
  model_full <- lm(mpg ~ ., data=bootstrap.sample)
  model_boot <- stepAIC(model_full, trace=FALSE)
  coeff_wt[i] <- model_boot$coef["wt"]
}

# Where the coefficient is NA it means the wt variable wasn't selected
# Set NA's to 0
coeff_wt[is.na(coeff_wt)] <- 0

# Histogram
Bootstrap <- data.frame(Coefficient=coeff_wt)
ggplot(Bootstrap, aes(x=Coefficient)) +
geom_histogram(aes(y = ..density..),binwidth = .5)+
  geom_density()+
ggtitle("Bootstrapped Coefficients")+
theme_bw()

# 95% bootstrap confidence interval
quantile(coeff_wt,c(.025,0.975))

```


