---
title: "Simple linear regression"
author: "Name"
date: "Date"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(performance)
library(dplyr)
library(corrplot)
```

# A simple linear regression using the cars dataset

'cars' is a built-in dataset in R: there is no need to load it.

## Summary statistics

```{r}
summary(cars)
```


## Scatter plot

Plot stopping distance (dist) against speed.

```{r}

library(ggplot2)
ggplot(cars, aes(x=speed, y=dist)) +
  geom_point() # + geom_smooth(method="lm")

# What does adding geom_smooth do?
```

What does this tell you?

Does this seem to be a good candidate for a simple linear regression?


## Correlation

Perform a linear correlation between these two variables.  Is the linear correlation statistically significant?

```{r}

with(cars, cor(speed,dist))
with(cars, cor.test(speed,dist))

```


## Linear regression

Model the stopping distance of these cars (from the 1920s). You will need to identify your dependent and independent variables.

```{r}

model1 <- lm(dist ~ speed, data=cars)
summary(model1)

```

Interpret the output of the model.

How would you explain the regression equation to someone?


$$\mbox{distance} = \beta_0 + \beta_1 \times \mbox{speed}$$

An increase of 1mph in speed leads to an average increase of beta_1 feet in distance.

How would you explain the R-squared value to someone?

The R-squared value tells us that XX% of the variation in distance can be explained by variation in speed and the fitted linear model.


## Model fit

Assess the performance of the model.

Does it show evidence of breaking model assumptions?

```{r}

library(performance)
check_model(model1)

```





# Dewis exercise 'r regression'

Use the code chunk below to contain your code to answer the Dewis question 

```{r Dewis}

# Start your code here


```


# The mtcars dataset

'mtcars' is a built-in dataset in R: there is no need to load it.

```{r}

library(dplyr)
df <- mtcars %>%
select(-c(am, vs, cyl, gear, carb))
glimpse(df)

```

Pick which variable is the best single variable for explaining the variation in miles per gallon (mpg), using a simple linear regression model.

Fully document your process, including your exploratory data analysis, model evaluation and selection.

How would you explain your chosen model to someone without technical expertise?

```{r}

## Summary statistics

## Correlation
### Hint: choose the variable that has the largest absolute correlation with miles per gallon

## Scatter plot

## Linear regression

## Interpretation

## Model fit

```





R packages mentioned today that you may find useful include:

- corrplot (for visualisation of correlations across a set of variables) <https://cran.r-project.org/web/packages/corrplot/vignettes/corrplot-intro.html>

- performance (for model evaluation and comparison) <https://easystats.github.io/see/articles/performance.html>


