---
title: "Relevant Codes for Exams"
author: "Dinka"
date: "2026-04-20"
output: html_document

# do not change this option, this shows the code in your knitted document
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```

# installing librabraries and packages 
``` {r}
if(!require()){install.packages("")} ## to check if package is required 

install.packages() ## installing a package, this can also be done using the tools in the status bar
library() ## checking if it is already installed 

```

# loading in data 
```{r}
content <- read.csv("file path")

```


``` {r str}
# data structure and variables
str() ## shows the structure of the data and variable types 
names() ## shows the columns of  a data frame
head() ## first few rows of a df 

## summary statistics 
``` {r}
summary() ## quick statistical overview, quartiles,mean median and mode
mean(mtcars$mpg) ## mean of a particular vector/ colunm '$' is the operator
round(mean(mtcars$mpg),2) ## rounding the mean to 2 decimal places 

```
 here , key insights or results of the summary will be outlined 
 
```{r eda 1}
 #Exploratory Data Analysis
 # Summary of numeric columns
summary(content[, sapply(content, is.numeric)], na.rm = TRUE)

# Frequency tables
table(content$handed)

table(content$handed, content$active)

# Mosaic plot
mosaicplot(table(content$handed, content$active))

# Summary statistics by handedness
by(content, factor(content$handed), summary)

# Proportions of handedness
prop.table(table(content$handed))
```
 outline key insights here 
 
```{r eda 2}
# visualization/ graphical summary 
## boxplots
ggplot(mtcars, aes(x = "", y = mpg)) +
  geom_boxplot(fill = "skyblue", color = "blue") +
  labs(title = "Boxplot of MPG from mtcars Dataset",
       y = "Miles per Gallon (mpg)") +
  theme_minimal()
```
 key insights here   
  
    
 ##  In this code:

- ggplot(mtcars, aes(x = "", y = mpg)): This sets up the basic ggplot object, specifying the dataset (mtcars) and aesthetics (mapping the empty string "" to the x-axis and the 'mpg' variable to the y-axis).

- geom_boxplot(fill = "skyblue", color = "blue"): This adds the boxplot layer to the ggplot object, specifying the fill and border color.

- labs(title = "Boxplot of MPG from mtcars Dataset", y = "Miles per Gallon (mpg)"): This sets the plot title and y-axis label.

- theme_minimal(): This applies a minimal theme to the plot for better aesthetics.


```{r eda 3}
## histogram
ggplot(mtcars,aes(mpg))+geom_histogram(binwidth = 4)+xlab("Miles per Gallon")+ylab("Number of Cars")+ggtitle("Distribution of Cars by Mileage")

##Editing this to show the histogram for the number of cylinders (edit the binwidth to be equal to 1).
ggplot(mtcars,aes(cyl))+geom_histogram(binwidth = 1)+xlab("Cylinders")+ylab("Number of Cars")+ggtitle("Distribution of Cars by Cylinders")
```



```{r data cleaning}


# Replace "unknown" with NA in 'handed'
content$handed[content$handed == 'unknown'] <- NA

# Replace extreme weight values with NA
content$weight[content$weight > 250] <- NA

# Count of non-missing weight values
length(na.omit(content$weight))

# Aggregated height statistics by handedness and activity
agg<- aggregate(
  height ~ handed + active,
  data = content,
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
```
here , key insights or results of the summary will be outlined 

```{r}
# correlation(between two variables)
with(cars, cor(speed,dist)) ## gives only the number(pearson's coefficient)
with(cars, cor.test(speed,dist)) ## gives other info like p-value, confidence interval etc.

```

```{r}
## linear regression(the dependent and independent variable needs to be identified )
model1 <- lm(dist ~ speed, data=cars)
summary(model1)

```
here you will Interpret the output of the model using the formula .
$$\mbox{distance} = \beta_0 + \beta_1 \times \mbox{speed}$$

```{r}
# this code removes am, vs, cyl, gear, carb

library(dplyr)
df <- mtcars %>%
select(-c(am, vs, cyl, gear, carb))
glimpse(df)

```
## Model fit

Assess the performance of the model.
```{r}

library(performance)
check_model(model1)

```



