---
title: "Week 4"
author: "Ben Derrick"
date: "2023-02-21"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Note: this worksheet uses the 'iris' dataset that exists within R. Only basic statistics and graphics are given in this worksheet to keep the focus on the hypothesis tests. 


Research Question 1. Is the population Sepal Width 3cm?
---------------------------------------------------------------

First, we write the null hypothesis formally.
$H_0: \mu = 3$


### EDA

I can get a quick summary of this variable (five number summary, mean as well as count of NA values):

```{r systolic_summary, echo=FALSE}
summary(iris$Sepal.Width)

```

One other key piece of EDA will be a histogram.

```{r hist_systolic, echo=FALSE}
hist(iris$Sepal.Width)
```

What does the EDA tell us about the truth (or otherwise) of the null hypothesis?


## A text book hypothesis test

The textbook hypothesis test in this case would be the "one-sample t-test".


### Assumptions
- no gross outliers
- observations independent
- sample large enough that $\bar{x}$ could be Normally distributed


### The test

```{r one_sample_t, echo=TRUE}
t.test(iris$Sepal.Width, mu = 3, alternative = "two.sided")
```


### The conclusion

What is the statistical conclusion to this hypothesis test?
(Remember this will be a statement about the null hypothesis.)

Application of the one sample t-test indicates that there is no evidence to suggest that the mean sepal width of the iris dlowers in not equal to 3 (p=0.1093)

=== A claim that the mean sepal width is equal to 3 is reasonable.

No violation of the assumptions apparent, histogram shows a pproximate normal distribution and no extreme outliers.

### The interpretation

1. Are the assumptions met (look at the histogram and the NA count)?
2. If we reject the null hypothesis, is the difference practically important?





Research Question 2. Do setosa have the same Sepal Widths as versicolor?
-----------------------------------------------------------------------------

Firstly we will filter so that only setos and versicolor species are included.


```{r means_by_sex}
Filtered_iris = subset (iris, Species == "setosa" | Species =="versicolor")
```


The null hypothesis in this case can be written down.

$H_0: \mu_S = \mu_V$

where $\mu_S$ denotes the Population mean value for setos, and $\mu_V$ denotes the Population mean value for versicolor.


### EDA

We can obtain group means

```{r means}
summary(Filtered_iris$Sepal.Width[iris$Species=="versicolor"])
summary(Filtered_iris$Sepal.Width[iris$Species=="setosa"])
```



And the most important visual will be a boxplot (or a faceted histogram).

```{r boxplot_Species}
boxplot(Filtered_iris$Sepal.Width ~ Filtered_iris$Species)
```

What does the EDA tell us about the truth (or otherwise) of the null hypothesis?


## Another text book hypothesis test

In this case we might use a two-sample t-test (independent samples).


### Assumptions

- no gross outliers, 
- observations independent, 
- sample large enough that $\bar{x_M}$ and $\bar{x_F}$ could be Normally distributed, 
- both groups have the same variance


### The test

```{r two_sample_t, echo=TRUE}
t.test(Filtered_iris$Sepal.Width ~ Filtered_iris$Species, var.equal=TRUE)
```


### The conclusion

What is the statistical conclusion to this hypothesis test?
(Remember this will be a statement about the null hypothesis.)


### The interpretation

1. Are the assumptions met (look at the histogram and the NA count)?
2. If we reject the null hypothesis, is the difference practically important?



Research Question 3. Is there a correlation between Sepal Length and Sepal Width?
-----------------------------------------------------------------------------

Note this one carefully, the null hypothesis is that the Population correlation coefficient is zero:

$H_0: \rho = 0$


### EDA



```{r scatter}
plot(iris$Sepal.Length,iris$Sepal.Width)

```

```{r correlation}
with(iris, cor(Sepal.Length, Sepal.Width, use="pairwise.complete.obs"))
```




## Hypothesis test

A hypothesis test assumes that Pearson's product moment correlation coefficient follows a t distribution with $n-2$ degrees of freedom if the samples follow independent normal distributions.  What other EDA might we need to check these assumptions?

### The test

```{r correlation_test}
with(iris, cor.test(Sepal.Length, Sepal.Width, use="pairwise.complete.obs"))
```

### The conclusion

What is the statistical conclusion to this hypothesis test?
(Remember this will be a statement about the null hypothesis.)


### The interpretation

1. Are the assumptions met - do you need some more EDA to check this
2. If we reject the null hypothesis, does this tell us that the correlation is practically important?




## Advanced activity using iris

Research Question 4. Do the three Species differ on Sepal Width
----------------------------------------------------------------------------------------------

This is an extension of the comparison of means we did for two Species, only now we have more than 2 groups and must use a different test. The null hypothesis:

$H_0: \mu_S = \mu_V = \mu_Vi$

(where S denotes setosa, V denotes versicolor,  Vi denotes virginoca)


### EDA



```{r boxplot_3}
boxplot(iris$Sepal.Width ~iris$Species)

```


## Hypothesis test

Because we are comparing more than one population mean, first of all we conduct a so-called "Analysis of variance" (ANOVA) against the null hypothesis that all means are the same.  


### Assumptions

1. There are no gross outliers, 
2. the observations are independent, 
3. the variance is the same in each group and 
4. the group means are Normally distributed


### The test

We can perform a classical ANOVA using the `aov()` command in R. The `print()` and `summary()` methods give slightly different information on the ANOVA object you create.

```{r econact_aov}
test3 <- aov(Sepal.Width ~ Species, iris)
print(test3)
summary(test3)
```