---
title: "Week3"
output:
  word_document: default
  html_document: default
date: "July 2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
#install.packages("ggplot2")
library(ggplot2)

```

## BACKGROUND

In this activity you will consider the usefulness of some suggested R code, identify data issues, and produce descriptive statistics.

A researcher is interested in the heights and weights of a population of adults, and whether these differ according to whether the adult has an active lifestyle or not and whether adults are left or right-handed. In general, they are interested in any interesting findings that can be found by exploring the dataset.


## SUPPORT

There is a supporting video that introduces this exercise, that uses the same data structure but different sample values. This may help you understand the context.

[Watch this video on YouTube](https://www.youtube.com/watch?v=Rkcm9FRFTB8)



## LOADING THE DATA

Load the dataset here! Call the dataset 'content'.

This approach assumes that you have a directory structure with subfolders called 'data' and 'scripts'. The data should be in the 'data' folder, and the Rmarkdown file should be saved in the 'scripts' folder. This is standard industry good practice for file directory structure. It helps towards the principles of reproducible research if we all follow the same directory structure anyone can run your code.

```{r data}

content <- read.csv("../data/weight_height.csv")


```

## DATA STRUCTURE

```{r structure}
str(content)
```
The variables collected for a number of adults are:
weight – in kilograms
height – in centimetres
handed - if adult is ‘left’ or right’ handed
active - is adult considered to be active? ‘yes’ or ‘no’



## Helpful Code

Some useful code is given below. Try it out and see what it does.

```{r summary1}

summary(content)

```

```{r summary}

# Summary of numeric columns
summary(content[, sapply(content, is.numeric)], na.rm = TRUE)

```

```{r useful}
# Summary of numeric columns
#summary(content[, sapply(content, is.numeric)], na.rm = TRUE)

# Frequency tables
table(content$handed)

table(content$handed, content$active)

# Mosaic plot
mosaicplot(table(content$handed, content$active))

# Summary statistics by handedness
by(content, factor(content$handed), summary)

# Proportions of handedness
prop.table(table(content$handed))

# Replace "unknown" with NA in 'handed'
content$handed[content$handed == 'unknown'] <- NA

# Replace extreme weight values with NA
content$weight[content$weight > 250] <- NA

# Count of non-missing weight values
length(na.omit(content$weight))

# Aggregated height statistics by handedness and activity
agg<- aggregate(
  height ~ handed + active,
  data = content,
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
```

## Preliminary exploratory data analyses.

Adapting the above code, or otherwise, prepare some summary statistics for the attached data. 

Consider what this reveals about any issues with the raw data, and further data preparation that may be required prior to statistical analyses. List those issues here.

```{r eda1}
str(content)

```


```{r eda}
summary (content)

```
Here’s a concise summary of the key insights from your statistics:

Weight: Values are fairly consistent, centered around ~80, with no obvious outliers. Distribution looks reasonable.

Height: Most values are realistic (around 175–182), but there is a major outlier (1792.4), which is likely a data entry error and should be investigated or removed.

Active & Handed: Both are stored as character variables, indicating categorical data that should ideally be converted to factors for proper analysis.

```{r eda1.1}
#boxplot to visualise the outlier or anomalies in the height

ggplot(content, aes(y=height)) + geom_boxplot()+
  labs(
    y = "height",
    title= "Distribution of height"
  )+
  theme_minimal()

```

```{r eda1.2}
#boxplot to visualise the outlier or anomalies in the height

ggplot(content, aes(y=weight)) + geom_boxplot()+
  labs(
    y = "weight",
    title= "Distribution of weight"
  )+
  theme_minimal()

```
# Key insight

There are extreme outliers (around approximately 1000 and approximately 1790), which are clearly unrealistic.

These outliers stretch the scale, making the main data look compressed.

```{r frquency table}
# Frequency tables
table(content$handed)

```

```{r prop frequency table}

# proportional Frequency tables
prop.table(table(content$handed))

table(content$handed, content$active)

```
```{r }
#mosaic plot to visually represent handedness by active
mosaicplot(table(content$handed, content$active))

```
```{r nulls}

# Count of non-missing weight values
length(is.na(content$weight))

```
```{r }

# Count of non-missing weight values
length(is.na(content$height))

```
## Data Cleaning


Apply any further data preparation that may be required prior to statistical analyses, to ensure that only valid observations will be included in future analyses.

```{r cleaning}
# replace extreme weight values with NA
content$weight[content$weight < 25] <- NA

# # replace extreme height values with NA
content$height[content$height < 25] <- NA

# replace extreme height values with NA
content$height[content$height < 140 | content$height > 220] <- NA

# replace missing value
content$handed[content$handed == "right" ] <- NA

# handling low frequency groups
content$handed[content$handed == "unknown"] <- NA

#check total values of NA
sum(is.na(content))


```
```{r boxplots}
#boxplot to confirm the outlier or anomalies in the height have been removed

ggplot(content, aes(y=height)) + geom_boxplot()+
  labs(
    y = "height",
    title= "Distribution of height"
  )+
  theme_minimal()

```

```{r boxplot}

#boxplot to confirm the outlier or anomalies in the weight have been removed

ggplot(content, aes(y=weight)) + geom_boxplot()+
  labs(
    y = "weight",
    title= "Distribution of weight"
  )+
  theme_minimal()
```
## Exploratory data analyses using only the valid observations.

Produce summary statistics using only the valid observations. You may wish to consider for example:

How many adults in the sample are active? 

What is the best estimate for the proportion of adults in the population that are active? 

How many active adults in the sample are right-handed? 

How many valid observations are there for weight? 

What is the sample minimum weight? 

What is the sample maximum weight? 

What is the sample mean weight? 

What is the sample standard deviation for weight? 

What is the sample minimum weight of the active adults? 

What is the sample maximum weight of the active adults? 

What is the sample mean weight of the active adults? 

What is the sample standard deviation weight of the active adults?

```{r eda3}


```



## Conclusions

Offer some interpretation. Some examples of research questions you may want to answer are:

How do active adults compare to non-active adults in terms of weight and height? 

How does the average and variability of weight and height differ by handedness?




